COVID-19 related social media use slightly decreases well-being

### 2023-01-30 manuscript.tex
- thorough update to paper - shortened introduction - added additional exploratory analyses - e.g., analysis of between person vars

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,mask]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Analyzing the effects of COVID-19 related social media use on well-being},
  pdflang={en-EN},
  pdfkeywords={COVID-19, well-being, social media, news use, panel study.},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{COVID-19, well-being, social media, news use, panel study.}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{endnotes}
\setlength{\parskip}{0em}
\raggedbottom
\note{\clearpage}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Analyzing the effects of COVID-19 related social media use on well-being}
\author{Name blinded for review\textsuperscript{1}}
\date{}

\shorttitle{COVID-19 related social media use and well-being}

@@

\abstract{%
In times of crisis such as the Corona pandemic citizens need to stay informed about recent events, political decisions, or mandatory protection measures. To this end, many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect. In this preregistered study, I investigated whether using social media for COVID-19 related reasons affected psychological well-being. To answer this question I analyzed data from the Austrian Corona Panel Project, which consists of 3,018 participants. Well-being was measured at all 24 waves, and communication at five specific waves. I ran three random effects within between models, controlling for several stable and varying confounders. Results showed that the effects of COVID-19 related social media use on well-being were very small, arguably too small to matter. Fears that social media use during times of crisis critically impairs well-being are not supported.
}

@@

During the COVID-19 pandemic,
numerous events unfolded in quick succession and several open questions emerged.
How dangerous is the virus?
Is it spreading in my region?
How is it transmitted, and how can I protect myself?
Because for many it was (and at the time of writing still is) a matter of life or death, people aimed to stay informed regarding the latest developments.
Governments around the world implemented safety measures, such as wearing masks, keeping physical distance, or enforcing lockdowns.
In this extraordinary situation, many people heavily relied on media to attain relevant information, and especially social media were at an all time high (Statista, 2021).

Some people actually couldn't stop using social media to learn about COVID-19 related news.
A new phenomenon emerged, termed ``doomscrolling'':
Users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or looking after their children (Klein, 2021).
It was asked whether using social media for COVID-19 related reasons is helpful, or whether it creates an additional burden on mental health (Sandstrom, Buchanan, Aknin, \& Lotun, 2021).
These concerns seem justified:
A study with 6,233 people from Germany conducted during the pandemic found that ``{[}f{]}requency, duration and diversity of media exposure were positively associated with more symptoms of depression'' (Bendau et al., 2021, p. 283).

As a result, with this study I want to build on this research and investigate whether or not COVID-19 related social media use affected well-being during the pandemic.
To this end, I analyzed a large-scale panel study from the Austrian Corona Panel Project (Kittel et al., 2020).
The panel consists of 24 waves and has an overall sample size of 3,018 participants.
The panel study collected a large number of psychological and demographic variables.
I explicitly aimed to investigate the causal effects of COVID-19 related social media use on well-being.

@@

The underlying theories that guided the selection of variables for my analysis are the two-continua model of mental health (Greenspoon \& Saklofske, 2001) and the hierarchical taxonomy of computer-mediated communication (Meier \& Reinecke, 2020).
According to the two-continua model, mental health consists of (a) psychopathology and (b) well-being.
Well-being can be differentiated into subjective and psychological well-being (Diener, Lucas, \& Oishi, 2018).
Whereas subjective well-being emphasizes hedonic aspects such as happiness and joy, psychological well-being addresses eudaimonic aspects such as fulfillment and meaning.
Subjective well-being is primarily about achieving positive affect and avoiding negative affect.
One of the most prominent indicators of well-being is life satisfaction.
In my view, because it represents a general appraisal of one's life, life satisfaction is best thought of as a meta concept that combines psychological and subjective well-being.
Notably, life satisfaction is stable and fluctuates only little, whereas it's the exact opposite for affect (Dienlin \& Johannes, 2020).
To capture well-being in this study I thus build on life satisfaction, positive affect, and negative affect.
Together, this should provide an encompassing perspective on potential media effects.

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology.
First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); and sixth, the message (e.g., content) (Meier \& Reinecke, 2020).
Whereas the first four levels focus on the communication \emph{channel}, the last two address the communication \emph{type}.
To measure social media use for the consumption of COVID-19 related news and topics, I here employ both the channel and the type of communication perspective, which together should offer a nuanced understanding of communication.

First, I investigate how well-being is affected by different types of communication affect, namely active and passive use.
Defining what constitutes active and what passive use is not always clear, and different understandings are currently discussed (Ellison, Triẹû, Schoenebeck, Brewer, \& Israni, 2020; Meier \& Reinecke, 2020).
Reading is generally considered as passive and writing as active, while there are also specific behaviors such as such liking or sharing content that fall somewhere in-between (Meier \& Krause, 2022).
In this study, I hence distinguish (a) \emph{reading} (passive), (b) \emph{posting} (active), and (c) \emph{liking and sharing} COVID-19 related posts (both active and passive).
Second, I analyze how using the most prominent branded applications affects well-being, and whether this effect changes across applications.
Branded apps are separate entities with potentially divergent effects.
Twitter might have a different effect as compared to WhatsApp because of their respective affordances.
For example, Waterloo, Baumgartner, Peter, and Valkenburg (2018) found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram.
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube.
Worth noting, this study is not about \emph{general} social media use during times of COVID, but on social media use \emph{focused} on COVID-19 related content.
Examples of such media use include posting thoughts about the pandemic or retweeting COVID-19 related news.

\hypertarget{theorizing-social-media-effects-on-well-being}{%
\subsection{Theorizing Social Media Effects on Well-Being}\label{theorizing-social-media-effects-on-well-being}}

From a theoretical perspective, how could we explain whether COVID-19 related social media use affects well-being?
According to the set-point model of subjective well-being, well-being is surprisingly stable (Lykken, 1999).
Although specific events such as marriage or salary can have significant effects, after some time well-being routinely returns to prior levels, which are mostly determined genetically (Sheldon \& Lucas, 2014).
Only very specific events and factors such as unemployment, disability, or death can cause long-term changes in well-being (Lucas, 2007).

Can media use be such a factor?
In advance, there doesn't seem to be a clear winner, and it's likely that both positive and negative effects cancel each other out.
Empirically, social media use on average oes not have a strong effect on well-being (Meier \& Reinecke, 2020).
According to the Differential Susceptibility to Media Effects Model (Valkenburg \& Peter, 2013), the effects of media use differ across individuals.
Whereas for some media are beneficial, for others they are harmful.
On average, however, effects are often small or negligible.
But also when focusing on individual users, social media can have both positive and negative effects on well-being (Büchi, 2021).
They can impair well-being when causing embarrassment, stress, or disinformation, and they can improve well-being when providing connectedness, information, or entertainment (Büchi, 2021).

Two of the most prominent media effect theories argue against strong negative impacts.
First, uses and gratifications theory states that people explicitly and rationally chose specific media because of their respective benefits (Katz, Blumler, \& Gurevitch, 1973).
If those benefits don't exist, they will spend their time elsewhere.
And social media offer copiuos benefits.
Most prominently, they help find relevant information, maintain and foster relationships, express one's personality, and entertain oneself (Pelletier, Krallman, Adams, \& Hancock, 2020).
The second major theory that needs to be listed here is mood management theory (Zillmann, 1988).
Users implicitly learn what media help them balance their mood and affect.
For example, when bored people use social media to entertain themselves.
Using experience sampling of well-being and logs of social media use, a study with 82 participants from Italy found that after episodes of social media use, levels of positive affect increased significantly (Marciano, Driver, Schulz, \& Camerini, 2022).

But people can also misjudge media affects and are often overly optimistic (Metzger \& Suh, 2017).
And precisely because social media have so many positive consequences, one can ask if this is not where the actual problem lies.
In other words, social media aren't problematic because they're inherently bad, but rather because they're too good.
And as with many other things, there can be too much of a good thing.
It is therefore often asked whether social media are addictive, and users sometimes express this fear themselves (Yang, Griffiths, Yan, \& Xu, 2021).
However, a recently published meta-analysis found that the two most prominent measures of addiction, the Bergen Facebook Addiction Scale and the Bergen Social Media Addiction Scale, have only small relations to well-being (Duradoni, Innocenti, \& Guazzini, 2020).
In addition, the general idea of labeling excessive social and new media use as addiction was criticized, arguing that social media represent new regular behaviors that should not be pathologized (Galer, 2018; van Rooij et al., 2018).

Because media effects can differ across situations (Valkenburg \& Peter, 2013), I now briefly focus on the effects of COVID-19 related social media use specifically.
First, one could assume a \emph{direct} negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Consuming such news can be depressing (Dörnemann, Boenisch, Schommer, Winkelhorst, \& Wingen, 2021), perhaps even changing some general perspectives on life.
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects (Dörnemann et al., 2021).
A study with 2.057 respondents from Italy reported that during the pandemic virtual community and social connectedness even increased during the pandemic (Guazzini, Pesce, Marotta, \& Duradoni, 2022).

There could also be \emph{indirect} effects.
When browsing social media for Covid-19 related news, many users reported being captivated to such an extent that they could not stop using social media (Klein, 2021).
During the pandemic social media use was at an all-time high in the US (Statista, 2021).
Although it is most likely that moderate social media use is not detrimental (Orben, 2020), overuse, however, might be more critical, and several studies have shown more pronounced negative effects for extreme users (Przybylski \& Weinstein, 2017).
To explain, overuse could impair well-being if it replaces meaningful or functional activities such as meeting others, working, actively relaxing, or exercising.
So if a society collectively overuses social media during a pandemic, there might be potential for negative effects.
On the other hand, one can make the case that overuse might even be beneficial, especially in times of a pandemic.
Exchanging COVID-19 related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is literally impossible at the time.
In situations where meaningful and functional activities are prohibited, using social media to exchange about COVID-19 related topics might not be the worst idea.
Besides, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand high quality information on current developments.

To summarize, from a theoretical perspective it is most likely that the average effects of social media use on well-being are negligible.
Building on established theories from Communication, we would not assume that effects are either profoundly negative or strongly positive.

\hypertarget{empirical-studies-on-social-media-effects}{%
\subsection{Empirical Studies on Social Media Effects}\label{empirical-studies-on-social-media-effects}}

So far, there is only little empirical research on how well-being is affected by COVID-19 related social media.
In their study on the relations between media use and mental health during the pandemic, Bendau et al. (2021) found that people who used social media as a primary source of information reported on average ``significantly more unspecific anxiety and depression {[}{]} and significantly more specific COVID-19 related anxiety symptoms'' (p.~288).
Eden, Johnson, Reinecke, and Grady (2020) analyzed the media use of 425 US college students during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, Choi and Choung (2021) reported that people who used media to attain information were more lonely and less satisfied with their lives.
Stainback, Hearne, and Trieu (2020) analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.
A four-wave panel study with 384 young adults from the U.S. analyzed the effects of general digital technology use---objectively measured via screenshots of screen-time applications---on mental health, separating within- and between-person relations (Sewall, Goldstein, \& Rosen, 2021).
The results showed that digital technology did not have any significant effects on mental health (for a similar study with comparable results, see Bradley \& Howard, 2021).
Together, the literature is mixed, with a slight focus on the negative effects of social media as news use (see also Dörnemann et al., 2021; Liu \& Tong, 2020; Riehm et al., 2020).
However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below).
We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education.

The question of whether and how social media use affects well-being \emph{in general}, on the other hand, is well-researched.
This also holds true for the different types of communication such as active or passive use.
A meta review (i.e., an analysis of meta-analyses) found that the relation between social media use and well-being is likely in the negative spectrum but very small (Meier \& Reinecke, 2020)---potentially too small to matter.
What determines whether or not an effect is considered small or trivial?
As a starting point, we could refer to standardized effect sizes.
According to Cohen (1992), small effect sizes start at \emph{r} = .10.
And indeed, several if not most of the current meta-analyses find effect sizes below that threshold (Ferguson et al., 2021; Huang, 2017; Meier \& Reinecke, 2020).

Finally, also several individual studies employing advanced methods found smalls relations between social media use and well-being (Keresteš \& Štulhofer, 2020; Orben, Dienlin, \& Przybylski, 2019; Przybylski, Nguyen, Law, \& Weinstein, 2021; Schemer, Masur, Geiß, Müller, \& Schäfer, 2021).
For example, Beyens, Pouwels, van Driel, Keijsers, and Valkenburg (2021) reported that although for some users (roughly one quarter) the effects of social media use on well-being were negative, for almost the same number of users they were positive, while for the rest the effects were neutral.
This finding is aligned the Differential Susceptibility to Media Effects Model:
Although there is substantial \emph{variation} of media effects for individual users, the \emph{average} effects reported in the literature are often small (Valkenburg \& Peter, 2013).

In conclusion, in light of the theoretical considerations and empirical studies presented above, I expect that COVID-19 related communication on social media doesn't affect well-being in a meaningful or relevant way.

@@

\hypertarget{current-study}{%
\section{Current Study}\label{current-study}}

@@

Testing this hypothesis, however, is not trivial.
First, in contrast to most hypotheses typically posited in the social sciences it implicitly contains an effect size, a so-called smallest effect size of interest (SESOI).
Effectively testing this hypothesis necessitates defining what's considered a ``trivial effect size'' and what's not.
Above I already referred to standardized effect sizes.
However, standardized effect sizes should only be a first step toward evaluating an effect's relevance (Baguley, 2009).
Standardized effect sizes are determined by a sample's variance,\footnote{Consider the effect size Cohen's \emph{d}: The mean's of the two groups that are to be compared are subtracted from one another and then divided by the sample's standard deviation (Cohen, 1992). Hence, if there is more deviation/variance in a sample, the effect size decreases, even if the difference of the group's means stays the same.} which is problematic:
The question of whether or not social media use affects a particular person in a relevant way should not depend on the variance in the sample in which that person's data were collected.
Instead, it should depend on absolute criteria.

What could be a minimally interesting, nontrivial effect?
Because this is a normative and ultimately philosophical question, there can never be a clear, single, or unanimous answer.
However, it is still necessary and helpful to try provide such a plausible benchmark.
I therefore suggest the following SESOI for this research question:

@@

In what follows, I list all the variables that I analyzed.
For the variables' means, range, and variance, see Table \ref{tab:tab-descriptives}.
For a complete list of all items and item characteristics, see \href{https://XMtRA.github.io/Austrian_Corona_Panel_Project}{companion website}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-mcelreathYesterdayClass2021}{}}%
McElreath, R. (2021). Yesterday in class, ... {[}Tweet{]}.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,mask,floatsintext]{apa7}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{COVID-19, well-being, social media, news use, panel study.}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother
\usepackage{endnotes}
\setlength{\parskip}{0em}
\raggedbottom
\note{\clearpage}

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={COVID-19 related social media use slightly decreases well-being},
  pdflang={en-EN},
  pdfkeywords={COVID-19, well-being, social media, news use, panel study.},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{COVID-19 related social media use slightly decreases well-being}
\author{Name blinded for review\textsuperscript{1}}
\date{}

\shorttitle{COVID-19 related social media use \& well-being}

@@

\abstract{%
In times of crisis such as the COVID-19 pandemic, citizens need to stay informed about recent political events. To this end, people increasingly use social media. However, being particularly engaging, many find it hard to disconnect from social media, especially during times of crisis. Using data from the Austrian Corona Panel Project consisting of 3,485 participants from 32 waves, controlling for several stable and varying confounders, the results showed that COVID-19 related social media use reduced facets of well-being. However, the effects were very small. Other factors such as health, income, exercise, or internal locus of control showed larger and more meaningful effects.
}

@@

During the COVID-19 pandemic it was critical to stay informed regarding the latest developments.
How dangerous is the virus?
In what region is it spreading?
How is it transmitted? What are the current safety regulations?
To obtain relevant information, many people heavily relied on social media, which were at an all time high (Statista, 2021).

Some actually could not stop using social media to learn about COVID-19 related news.
A new phenomenon termed ``doomscrolling'' emerged.
Many users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or even looking after their children (Klein, 2021).
In the media it was hence increasingly asked whether using social media for COVID-19 related reasons would, next to all other stressors, create an additional burden on mental health (Sandstrom et al., 2021).
Although research has begun addressing this question (Bendau et al., 2021; Bradley \& Howard, 2021; Choi \& Choung, 2021; Dörnemann et al., 2021; Eden et al., 2020; Guazzini et al., 2022; Latikka et al., 2022; Liu \& Tong, 2020; Riehm et al., 2020; Sewall et al., 2021; Stainback et al., 2020; Yue et al., 2022), it still largely unknown if COVID-19 related social media use during the pandemic has had a meaningful impact on well-being.
By analyzing large-scale panel data from Austria and by focusing on intra-individual causal effects, this study aims to answer to this question.

@@

This study investigates how different \emph{facets} of well-being are affected by different \emph{types} and different \emph{channels} of communication (Meier \& Reinecke, 2020).
Building on the typology of subjective well-being (Diener et al., 2018), three different well-being facets are analyzed: life satisfaction, positive affect, and negative affect.
Because effects of social media depend on how they are used (Verduyn et al., 2015), I further distinguish three types of use and five popular channels.
The types of use include reading, posting, and liking and sharing COVID-19 related posts.
In doing so, this study analyzes social media use focused on COVID-19 related content, which includes posting thoughts about the pandemic, reading posts and comments, or retweeting COVID-19 related news.
The five channels to be investigated are Facebook, Twitter, Instagram, WhatsApp, and YouTube, which at the time ranked among the most popular social media services in Austria.

How could the various types and channels of COVID-19 related social media use affect well-being?
According to the set-point theory, well-being is surprisingly stable (Lykken, 1999).
Although specific events such as marriage or salary can have significant impacts on well-being, in most cases effects are only short-term, with well-being after some time routinely returning to prior levels (Sheldon \& Lucas, 2014).
Only very specific factors such as unemployment, disability, or death can cause long-term changes in well-being (Lucas, 2007).
Can media use be such a factor?
Current literature overviews suggest no:
Although social media use on average decreases well-being, the effects are small (Meier \& Reinecke, 2020).
In addition, the effects of media use differ across individuals and types of content (Valkenburg \& Peter, 2013).
Whereas for some users social media are more beneficial, for others they are more harmful.
For example, in one study it was estimated that roughly one quarter of all users experienced negative effects, another quarter positive effects, while for the rest the effects were neutral (Beyens et al., 2021).
This finding is aligned with the Differential Susceptibility to Media Effects Model:
Although there is substantial \emph{variation} of media effects for individual users, there are both negative and positive aspects, leading to small effects on average (Valkenburg \& Peter, 2013)---as confirmed by several literature overviews (Dienlin \& Johannes, 2020; Huang, 2017; Meier \& Reinecke, 2020; Orben, 2020).

Why are the effects of social media use on well-being small on average?
Using Social media is ambivalent.
It can impair well-being when causing embarrassment, stress, or disinformation, and it can improve well-being when providing connectedness, information, or entertainment (Büchi, 2021).
Although social media are often associated with negative outcomes, there are a large number of positive effects, which either offset or at least significantly reduce negative outcomes.
Positive outcomes encompass finding relevant information; maintaining and fostering relationships; expressing one's personality; or entertaining oneself (e.g., Pelletier et al., 2020).
Negative effects include distraction; displacement of other meaningful activities; consumption of shallow, misleading or challenging content; or negative social comparisons (e.g., Meier \& Krause, 2022).
People are much more likely to use media that offer plentiful benefits (Katz et al., 1973).
Users implicitly learn what media help them regulate their mood and thereby well-being (Zillmann, 1988).
Given the ubiquitous use of social media, the many benefits they offer, plus the general and implicit competence to use media that foster mood management, it is hence unlikely that the effects of social media use on well-being are profusely negative.

There is still little empirical research on how well-being is affected by social media use that is focused on COVID-19 specifically.
Results are mixed.
When browsing social media for COVID-19 related news, many users reported being captivated to such an extent they could not stop using social media (Klein, 2021).
In a study with 1,131 residents from Wuhan in China, people who spent more time in quarantine also spent more time on social media (Yue et al., 2022).
Those who spent more time on social media also engaged in more upward social comparison, which was related to increased levels of stress.
People who used social media as a primary source of information reported on average ``significantly more unspecific anxiety and depression {[}\ldots{]} and significantly more specific COVID-19 related anxiety symptoms'' (Bendau et al., 2021, p. 288).
Eden et al. (2020) analyzed how 425 US college students used media during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a large-scale study with 11,537 respondents from the US, increased COVID-19-related media consumption was related to more psychological distress (Stainback et al., 2020).
A four-wave panel study with 384 young adults from the U.S. analyzed the effects of general digital technology use during the pandemic on mental health, finding that digital technology did not have significant effects on mental health (Sewall et al., 2021; for a similar study with comparable results, see Bradley \& Howard, 2021).
A study with 2.057 respondents from Italy reported that during the pandemic virtual community and social connectedness increased (Guazzini et al., 2022).
In a study with 735 participants from Finland, levels of loneliness were stable during the pandemic, and people who engaged more on social media experienced less loneliness (Latikka et al., 2022).
Together, the literature is mixed, showing both positive and negative affects of social media use focused on or during COVID-19 on well-being (see also Dörnemann et al., 2021; Liu \& Tong, 2020; Riehm et al., 2020).

In conclusion, these mixed empirical results, together with the observation that social media effects on well-being are very small in general, I expect that COVID-19 related communication on social media does not affect well-being in a meaningful or relevant way.

@@

The hypotheses, the sample, the measures, the analyses, and the inference criteria (SESOI, \emph{p}-value) were preregistered on the Open Science Framework, accessible here: \url{https://osf.io/87b24/?view_only=b2289b6fec214fa88ee75a18d45c18f3}.
Because in this study I analyze data from an already existing large-scale data set, the preregistration was done prior to accessing the data.
The preregistration was designed on the basis of the panel documentation online (Kittel et al., 2020).
In some cases, it was impossible to execute the analyses as I had originally planned, for example because some properties of the variables only became apparent when seeing the actual data.
The most relevant deviations are reported below, and a complete list of all changes can be found in the online \href{https://XMtRA.github.io/Austrian_Corona_Panel_Project}{companion website} (\url{https://XMtRA.github.io/Austrian_Corona_Panel_Project}).

The data come from the Austrian Corona Panel Project (Kittel et al., 2021), which is a large-scale standalone panel study.
The data are hosted on AUSSDA, are publicly available here (\url{https://doi.org/10.11587/28KQNS}), and consist of 32 waves.
The study was conducted between March 2020 and June 2022, and data collection is now finished.
Between March 2020 and July 2020, the intervals between waves were weekly and afterward monthly.
Each wave consists of at least 1,500 respondents.
The sample size was \emph{N} = 3,485, with overall 111,520 observations.
Panel mortality was compensated through a continuous re-acquisition of new participants.
Participants were sampled from a pre-existing online access panel provided by the company Marketagent, Austria.
Panel members were incentivized with 180 credit points for each wave of the study.

Achieved via quota sampling, the sample matched the Austrian population in terms of age, gender, region/state, municipality size, and educational level.
In order to participate in the study, the respondents needed to be Austrian residents and had to be at least 14 years of age.
All respondents needed to have access to the internet (via computer or mobile devices such as smartphones or tablets).
Ethical review and approval was not required for the study in accordance with the local legislation and institutional requirements.
The participants provided their written informed consent to participate in this study.
The average age was 41 years, 49 percent were male, 14 percent had a University degree, and 5 percent were currently unemployed.

@@

Testing the hypothesis necessitates defining what is considered a ``trivial effect size''.
Being a normative question, finding a clear, single, or unanimous answer is impossible.
However, it is still necessary and helpful to work toward a plausible benchmark.
I suggest the following SESOI for this research question:

@@

For the variables' means, range, and variance, see Table \ref{tab:tab-descriptives}.
For a complete list of all items and item characteristics, see \href{https://XMtRA.github.io/Austrian_Corona_Panel_Project}{companion website}.

@@
